### 2020-02-17 ordinal/ordinal.Rmd
ordinal tutorial

```{r init, include=F}
knitr() ## Zdefiniowa³em w .Rprofile, ustawia domy¶lne opcje chunków

sdt_ml_plot = function(model, adata, sdata){
    sdt = !is.null(adata$stimulus)
    o = optimizing(stan_model(model_code = model), data = sdata, init = 0)
    mp = matrix(o$par[grep('multinomial_p', names(o$par))], nrow = nrow(adata$data))
    obs = t(apply(adata$counts, 1, function(x)(x / sum(x))))
    if(sdt)
        stim = as.factor(adata$stimulus)
    f = as.factor(adata$data[,1])
    if(ncol(adata$data) > 1)
        for(i in 2:ncol(adata$data))
            f = f:as.factor(adata$data[,i])
    if(sdt){
        ggplot(data.frame(f = rep(f, ncol(mp)), stim = rep(stim, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs, group = stim, color = stim)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'Subj x Condition panels, lines = model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }else{
        ggplot(data.frame(f = rep(f, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'lines = fitted model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }
}
simulate_from_fit = function(fit, adata, fixed, random){
    ## Realistic known parameter values
    s = apply(as.data.frame(fit), 2, mean)
    sdata = make_stan_data(adata, fixed, random)
    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
    for(r in 1:nrow(gamma_fixed))
        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
    gamma_random = list()
    for(r in 1:sdata$group_1_size)
        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
    delta_fixed = s[grep('delta_fixed', names(s))]
    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
    for(r in 1:nrow(delta_random))
        for(i in 1:ncol(delta_random))
            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d,1]', r, i)]
    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
    dprim = rep(NA, sdata$N)
    for(n in 1:sdata$N){
        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
    }
    criteria = t(apply(gamma, 1, function(x){
        exp_x_0 = exp(c(x, 0))
        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
    }))
    distr_shift = -sdata$stim_sign * dprim / 2
    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
    for(k in 2:(sdata$K - 1)){
        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
    }
    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
    data_sim = adata
    ## Now we can simulate
    for(r in 1:nrow(adata$counts))
        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
    data_sim$dprim = dprim
    data_sim$gamma = gamma
    data_sim$criteria = criteria
    data_sim
}

## Works only when there is just one grouping factor
simulate_from_fit = function(fit, adata, fixed, random, gamma_link){
    s = apply(as.data.frame(fit), 2, mean)
    sdata = make_stan_data(adata, fixed, random)
    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
    for(r in 1:nrow(gamma_fixed))
        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
    gamma_random = list()
    for(r in 1:sdata$group_1_size)
        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
    delta_fixed = s[grep('delta_fixed\\[', names(s))]
    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
    for(r in 1:nrow(delta_random))
        for(i in 1:ncol(delta_random))
            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d]', r, i)]
    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
    dprim = rep(NA, sdata$N)
    for(n in 1:sdata$N){
        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
    }
    criteria = t(apply(gamma, 1, function(x){
        exp_x_0 = exp(c(x, 0))
        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
    }))
    distr_shift = -sdata$stim_sign * dprim / 2
    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
    for(k in 2:(sdata$K - 1)){
        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
    }
    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
    data_sim = adata
    ## Now we can simulate
    for(r in 1:nrow(adata$counts))
        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
    data_sim$dprim = dprim
    data_sim$gamma = gamma
    data_sim$criteria = criteria
    data_sim
}
```

It is now possible to fit hierarchical ordinal polytomous regression
models in bhsdtr by using order-preserving link functions for the
thresholds. This way not only the constant shift in the thresholds,
but also the differences in the pattern of the thresholds can be
accounted for in certain situations. To my knowledge this is a new
kind of hierarchical ordinal model that can be used to estimate
ordinal polytomous item-parameter variability in a way that until now
was not possible. If you have a bunch of Likert-type items, or
confidence ratings, or anything of this sort in your data, than you
may find this model useful.

Because by definition a polytomous item has a multinomial distribution
and the multinomial distribution does not have a separate parameter
for the residual variance, when item-parameter variability is not
accounted for by the model it can only be misplaced
(overdispersion). The correct way to account for item-parameter
variability in ordinal polytomous items is to allow for individual
threshold effects while respecting the ordering of the
thresholds. This can only be achieved by using the order-preserving
link functions.

Assume that the ordered responses ranging from 1 to K come from the
same general kind of process as the responses in an SDT model when the
stimulus is 1 ("noise"). For example, when John is faced with the
Likert-type item "How often did you cry during the last two weeks" an
unknown process produces some "internal experience" value. This latent
value manifests itself in the response "never" (= 1) if it is low
enough, "a few times" (= 2) if it is a bit higher than some latent
threshold, or in the response "frequently" (= 3) if it is higher than
some other, still higher, threshold. The thresholds are assumed to be
ordered, because otherwise the item would not have the intended
meaning and thus would not be a valid instrument of measuring
depression.

Now consider a participant who repeatedly responds to the same item
and who has the memory of this event wiped out after each
response. Despite the wiping out of memory we can still expect some
variability in the responses due to the instability of the complex
process which produces the internal value, as well as due to the
instability of the decision criteria. This variability is certainly
there, the question is if it is non-negligible, and if so, what can be
done about it. Unfortunately, without additional assumptions it is
impossible to model both kinds of variability at the same time even in
the hypothetical repeated measurement with memory wipeout scenario,
because every shift in the internal value is indistiguishible from a
uniform effect in the thresholds, i.e., a constant shift of all the
thresholds. As I will explain shortly, the required additional
assumptions that make such models identifiable under certain
conditions are necessarily *causal*.

In many situations this is both a statistical and a conceptual
problem. In particular, when the origin of the internal value is
unknown because the design is purely observational (e.g.,
questionnaire measurement), the way of describing the process of
responding to a Likert-type item in terms of internal values and
thresholds is simply redundant. As far as the design of the study is
concerned, the value of the internal evidence "comes from nowhere" -
all we can say about it is that it is relative to the thresholds - and
the position of each threshold is relative to the positions of other
thresholds and to the position of the internal value. In this case to
say that John thinks that he was not crying at all during the last two
weeks and to say that *John's* threshold for the response "a few
times" is above the amount of crying that John thinks he experienced,
or decided to reveal, is to say the same thing.

It follows that *in such situations we do not need the internal value
parameter at all*, unless there is some factor about which we can
perhaps assume that it can *only* produce a uniform effect in the
thresholds, such as the stimulus class ("noise" or "signal") in a
binary classification task, which can *perhaps* be assumed to induce a
change (*d'*) in the internal value only; In such cases the internal
value parameter may be justified and useful, because it does constrain
the model: in the repeated measurement case the model is no longer
saturated.

But what about the promise of Item Response Theory? Isn't it true that
we can separate the internal values (so called "person parameters")
from the thresholds (so called "item parameters") and place them on a
common scale? Once the participants leave the room the items retain
their estimated "difficulty levels", right? The item difficulty
(=thresholds) is not defined relative to some sample of participants
who filled the questionnaire. How is this possible if the two kinds of
effects are completely confounded? In particular, how does Item
Response Theory achieve when each person provides only a single
response to each item?

It is not an *achievment* of Item Response Theory, but it's
(unrealistic) *assumption*. The illusory nature of this achievement
becomes obvious as soon as we realize that without additional
(causal!)  assumptions it is impossible to distinguish between genuine
internal value effects and (dis)simulation (i.e., faking), which
conteptually corresponds to the change in the *participant*-specific,
not item-specific, decision thresholds, also known as
"item-parameters". There is nothing in the *desing* of the process of
measurement by a questionnaire that makes it possible to distinguish
genuine differences in internal values from the differences in the way
of responding. This "achievement" of Item Response Theory is a
consequence of treating the "item-parameters" as constant accross
participants when fitting an IRT model.

Because of random assignment of the stimulus class above chance
performance in a binary classification task cannot be faked, and so
the *d'* paramater definately captures at least part of the effect of
the stimulus on the internal value. On the other hand, every possible
pattern of responses to a questionnaire can be faked. The only way to
distinguish the "item" and "person" effects is to *assume selective
influence*, i.e., that there is some factor f (such as the stimulus
class in a binary classification task, or, in case of a depression
questionnaire, participants trait, or a therapeutic intervention)
which can *only* affect the internal value, thus producing a
*constant* shift in the thresholds. The assumption of selective
influence makes the model with the two kinds of parameters
identifiable if there are repeated measurements for at least one
participant and one item or a large number of participants provides
single responses to all of the items. In particular, in typical
applications of IRT it is simply *assumed* in the model that the
participants differ in the internal values, and the item parameters
are constant accross the sample of participants (selectivity).

In every book on Item Response Theory that I know of the issue of
measurement invariance due to the variability in the "item-parameters"
is reduced to the problem of Differential Item Functioning
(DIF). However, the variability in "item-parameters" that I talk about
here has nothing to do with DIF, because DIF is a *population* level
effect; An estimate of DIF is obtained when the estimates of
"item-parameters" obtained for two different (large) samples of
participants are compared. However, these estimates are obtained by
fixing the "item-parameters" within each sample. It does not take much
thought to see that the assumption of constant "item-parameters" at
the group/population/sample level is utterly unrealistic. It follows
that the achievement of separating participant and item effects in IRT
models is not an achievement at all, it is just wishful thinking
combined with aggressive marketing.

Some of the problems related to item-parameter variability can be
addressed *when additional causal assumptions of selective influence
are introduced and the design of the study makes the model
identifiable*. For example, we can use the data from a binary
classification task and model only the responses for one kind of
stimulus, e.g., "noise".

Here is a simple example which illustrates how we can fit a
hierarchical ordinal polytomous regression model using the bhsdtr
package. We will be using the data from one condition in the gabor
dataset:

```{r c1, echo=T}
library(ggplot2)
library(bhsdtr)
library(Hmisc)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left,
## r = 5 ~ no experience right, ... r = 8 ~ absolutely clear image right
gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
d1 = gabor[gabor$duration == '32 ms' & gabor$order == 'DECISION-RATING',]
## Only the left-tilted gabor patches
d1 = d1[d1$stim == 0,]
```

To fit a hierachical model we have to aggregate the data (this makes
the process of sampling much faster), create the fixed- and
random-effects lists, and create the data structures and the model
code required by stan. Note that we have introduced a new parameter -
eta - which represents the internal value. As we already know, without
the selectivity assumption the difference between item and person
parameters is arbitrary and so in bhsdtr the first element of the
fixed effects vector for the eta parameter is automatically fixed at 0
eto make the model identifiable (this bahavior may change in the near
future):

```{r c2, echo = T}        
## This order-preserving link function represents the main criterion
## directly as a free paramter and all the other criteria as log of
## distance between the criterion and the criterion adjacent to it
## which is closer to the main criterion
link = 'log_distance'
adata = aggregate_responses(d1, response = 'r', variables = c('id'))
fixed = list(eta = ~ 1, gamma = ~ 1)
random = list(list(group = ~ id, gamma = ~ 1))
model = make_stan_model(random, model = 'ordinal', gamma_link = link)
sdata = make_stan_data(adata, fixed, random, model = 'ordinal', gamma_link = link)
```

To simplify the task we will use maximum likelihood optimization,
since the model is saturated anyway:

```{r c3, echo=T}
o = optimizing(stan_model(model_code = model), sdata, init = 0)
```

Unsuprisingly, the model fits perfectly for every participant:

```{r p1}
sdt_ml_plot(model, adata, sdata)
```

Now we will add the estimated gamma/criteria random effects to the
fixed effects to obtain participant-specific gamma/criteria estimates,
convert between the gamma and the criteria vectors, and take a look at
the estimated variability in the criteria:

```{r c4,echo=T}
gamma_fixed = o$par[grep('gamma_fixed\\[', names(o$par))]
gamma_random = matrix(o$par[grep('gamma_random_1', names(o$par))], nrow = sdata$group_1_size)
for(i in 1:nrow(gamma_random))
    gamma_random[i,] = gamma_random[i,] + gamma_fixed
criteria = gamma_to_crit(gamma_random, beta_index = NULL, gamma_link = link)
df = expand.grid(criterion = 1:7, id = 1:nrow(criteria))
df$estimate = as.vector(t(criteria))
ggplot(df, aes(id, estimate, group = id)) + geom_line() + geom_point()
```
The thresholds may look a bit weird, but we have to rememberd that
what we see here is *a mixture of d' and threshold effects*. The
estimates are somewhat similar to what we would have obtained if we
fitted an SDT model, but they clearly do not have the intended
interpretation; They do not deconfound the two things that an SDT
model is supposed to deconfound, namely sensitivity and bias. Contrary
to what seems to be commonly assumed, if we included the stimulus
class we would not solve the problem - for that we have to also
introduce the strong causal assumption of selective influence, which,
like any other strong assumption, can easily be false.

The justification of the selective influence assumption comes from a
psychological theory of the task, not from the design of the
study. All that is guaranteed by randomly assigning the stimulus class
(i.e., by choosing the stimulus class at random on every trial) is
that any statistical effect of the stimulus class is an unbiased
estimate of the total causal effect of the stimulus class. In order to
empirically justify the assumption of *selective* influence of the
stimulus class on the internal evidence value we would have to *test*
if the stimulus class does not also affect the thresholds.

We can do exactly this. We will use the 'log_distance' link function,
because when this link function is used one of the elements of the
gamma vector directly represents the position of the main decision
criterion, and all the other elements represent the pattern of the
distances between the thresholds. If the stimulus does not affect the
pattern of the thresholds we should observe an effect only for the
main criterion parameter. This model is not exactly equivalent to a
hierarchical SDT model without selective influence since we do not
constrain the effect of the stimulus class on the main criterion to be
non-negative, but there is enough data points per participant that we
can at least hope that this is not a serious problem.

KOD

>>>>

```{r init, include=F,echo=F}
knitr::opts_chunk$set(dev = 'svg', comment = NA, fig.path='Figs/',fig.width=8, fig.height=6,
                      echo=TRUE, warning = FALSE, error = FALSE, message = FALSE,
                      cache.path = '~/windows/temp/ordinal', cache=T)

sdt_ml_plot = function(model, adata, sdata){
    sdt = !is.null(adata$stimulus)
    o = optimizing(stan_model(model_code = model), data = sdata, init = 0)
    mp = matrix(o$par[grep('multinomial_p', names(o$par))], nrow = nrow(adata$data))
    obs = t(apply(adata$counts, 1, function(x)(x / sum(x))))
    if(sdt)
        stim = as.factor(adata$stimulus)
    f = as.factor(adata$data[,1])
    if(ncol(adata$data) > 1)
        for(i in 2:ncol(adata$data))
            f = f:as.factor(adata$data[,i])
    if(sdt){
        ggplot(data.frame(f = rep(f, ncol(mp)), stim = rep(stim, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs, group = stim, color = stim)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'Subj x Condition panels, lines = model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }else{
        ggplot(data.frame(f = rep(f, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'lines = fitted model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }
}
```

# Hierarchical ordinal regression models in bhsdtr

This document contains a short tutorial on fitting hierarchical
ordinal polytomous regression models using the bhsdtr package, but
there is more; It also contains some explanations, illustrated with
annotated code, of several issues related to ordinal models in
general, and Signal Detection Theory and Item Response Theory in
particular. If you are interested in SDT, IRT, or ordinal models in
general, perhaps you will find something useful, new, or even mildly
provocative here.

Anyway, it is now possible to fit hierarchical (more than two possible
values) ordinal regression models in bhsdtr by using order-preserving
link functions for the thresholds. If you have a bunch of Likert-type
items, or confidence ratings, or PAS ratings, or anything of this sort
in your data, then this model may suit your needs. To my knowledge
this is a new kind of hierarchical ordinal model that can be used to
estimate variability in the thresholds in a way that until recently
was not possible.

Ordinal polytomous variables are common in psychology, they are
usually analyzed by fitting a linear model, and all sorts of bad
things happen because of this, but only a few people seem to care. I
for one do not always care about this. We will not address this kind
of problems here, because we will analyze ordinal data using
(hierarchical) ordinal models.

We will use the following libraries and global variables:

```{r echo=T,results="show"}
knitr::opts_chunk$set(echo=TRUE,results="show")
library(ggplot2)
library(bhsdtr)
library(Hmisc)
library(coda)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
## This variable indicates if the fitted models can be loaded from disk
fresh_start = FALSE
sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
```

First, we will consider an ordinal model which assumes that there is
an underlying normally distributed latent variable (e.g., an evidence
distribution in Signal Detection Theory, or a latent trait in an IRT
model) and when the value of this variable is above the latent
threshold *i*, but below the threshold *i + i*, the response *i + 1*
is generated. Threshold *0* is fixed at *-infinity* and threshold *K*
is fixed at *+infinity*, which means that there are *K-1* "regular"
thresholds and *K* possible responses. Since once we know the total
number of observations and the number of occurrences of *K - 1*
different responses, the number of occurrences of the only remaining
kind of response is known, there are *K - 1* degrees of freedom in
such data. The response variable is ordinal, and so the thresholds are
ordered. This is how this model looks like when the thresholds are
equidistant and centered at the mean of the latent variable
distribution (they do not have to be equidistant nor do they have to
be centered):

```{r echo=F}
x = seq(-3, 3, length.out = 1000)
y = dnorm(x)
p = ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_line(lty = 2)
crit = c(-2, -1, 0, 1, 2)
for(i in 1:length(crit))
    p = p + eval(substitute(geom_vline(aes(xintercept = cr)), list(cr = crit[i])))
p = p + scale_x_continuous(breaks = crit, labels = 1:length(crit), name = 'threshold / criterion')
for(i in 1:(length(crit) + 1))
    p = p + eval(substitute(geom_text(aes(x, y, label = label, size = 20)),
                            list(x = c(-3, crit)[i] + .5, y = max(y) / 2, label = paste(i))))
p
```

The number of parameters in this model matches the degrees of freedom
in the data, which means that the model is saturated, which means that
it always fits the data, but it does not mean that there is no
uncertainty associated with the estimates of its parameters.

We will be using this dataset...

```{r echo=T,results="show"}
data(gabor)
```

... from which we will select two subsets of data for one condition,
one subset (*d1*) with only the "noise" trials and the other (*d2*)
with both the "noise" and the "signal" stimuli. We will also calculate
the response variable by combining the information about the stimulus,
the accuracy, and the rating.

```{r echo=T,results="show"}
## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left, r
## = 5 ~ no experience right, ... r = 8 ~ absolutely clear image right
gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
d1 = gabor[(gabor$duration == '32 ms') & (gabor$order == 'DECISION-RATING') & (gabor$stim == 0),]
d2 = gabor[(gabor$duration == '32 ms') & (gabor$order == 'DECISION-RATING'),]
```

We cannot naively model the effects in the thresholds, because the
thresholds have to stay ordered, and assuming, for example, that each
threshold has a normal distribution due to the effect of participants
would violate the ordering assumption.

To supplement our model with a hierarchical linear regression
structure the vector of the thresholds has to be translated to the
unconstrained vector gamma. The 'log_distance' order-preserving link
function is one solution to this problem, but other order-preserving
link functions are also implemented in the bhsdtr package. The choice
of the link function may sometimes be important, but, since we will
not introduce numerical predictors for the thresholds, we can choose
any link function which seems most convenient.

The log_distance function is convenient for our purposes because it is
simple, and because the elements of the gamma vector under this link
function are easy to interpret and may be directly related to simple
research questions. In particular, one of the elements of the gamma
vector under the log_distance link function, i.e., the middle
threshold, will soon become important. This link function represents
the main threshold/criterion directly as a free parameter (here
gamma[4], because K = 8, and 8 / 2 = 4); Every other criterion is
represented as log of distance between the criterion and the one
criterion adjacent to it which is closer to the main criterion. This
way the order is preserved, and each threshold *i* is associated with
the corresponding unconstrained element *i* of the gamma
vector. Random and fixed effects in the thresholds are represented as
(hierarchical) linear effects in gamma. The two representations are
isomorphic, which means that we can always translate between them if
we need to.

```{r echo=T,results="show"}        
link = 'log_distance'
```

To fit a model we have to aggregate the data (this makes the process
of sampling much faster), create the fixed- and random-effects lists,
and create the data structures and the model code required by
stan. Note that here we introduce a new parameter - eta - which
represents the average internal value. If we are going to model all
the thresholds (gamma parameter vector) then this parameter is
redundant, because a constant shift in the thresholds is equivalent to
the opposite effect in eta. The bhsdtr package will automatically fix
the first element (here the only element) of the eta fixed effects
vector at 0.

First, we will fit a model which assumes that the pattern of the
thresholds is the same for all the participants, but the mean of the
latent distribution may differ across participants. If this makes
little sense to you please take a look at the picture of the ordinal
model with ordered thresholds and imagine that the underlying
distribution has a different mean for different participants.

Note that the two assumptions, i.e., that the thresholds are constant
across participants and latent means or samples are participant-, not
item-specific seem to be accepted in the great majority of
applications of Item Response Theory. This is reflected in the IRT
terminology: thresholds are often called "item-parameters" and latent
means or samples are often called "person-parameters". Some even say
that IRT can provide separate estimates of the true item- and
person-parameters.

```{r echo=T,results="show"}
fixed = list(eta = ~ 1, gamma = ~ 1)
random = list(list(group = ~ id, eta = ~ 1))
adata = aggregate_responses(d1, response = 'r', variables = 'id')
model = make_stan_model(random, model = 'ordinal', gamma_link = link)
sdata = make_stan_data(adata, fixed, random, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit = stan(model_code = model, data = sdata,
                   pars = c('eta_fixed', 'gamma_fixed', 'eta_random_1', 'eta_sd_1'),
                   init_r = .5,
                   iter = 5000,
                   chains = 4)
    saveRDS(fit, '~/windows/temp/ordinal_fit.rds')
}else{
    fit = readRDS('~/windows/temp/ordinal_fit.rds')
}
print(fit, probs = c(.025, .975), pars = c('eta_fixed', 'gamma_fixed'))
```

The neff and Rhat columns look good. Here is how the model fits (this
is a maximum likelihood fit):

```{r echo=T,results="show"}
sdt_ml_plot(model, adata, sdata)
```

So how do the "item-parameters" look like according to this model?

```{r echo=T,results="show"}
crit = gamma_to_crit(as.data.frame(fit))
round(t(apply(crit, 2, sumfun)), 2)
```

Note that the main criterion (gamma[4,1]) is *positive*, which makes
sense, since when the stimulus is of class "noise" the middle
criterion / threshold / average position of the thresholds is *to the
right* of the evidence distribution mean. Otherwise, as we will see
later, in this particular case, and perhaps also in many other similar
cases, these thresholds-as-"item-parameters" are probably meaningless.

Now we will introduce the model which accounts for the
inter-individual differences in the pattern of the thresholds by
modelling the gamma random effects. We cannot simultaneously model the
eta random effects in this model, since all the information that can
be captured by the eta parameter (i.e., the constant shift in the
thresholds) is already captured by the fixed and the random gamma effects.

```{r echo=T,results="show"}                
adata0 = aggregate_responses(d1, response = 'r', variables = c('id'))
## The only element of the eta fixed effects vector will be automatically fixed at 0
fixed0 = list(eta = ~ 1, gamma = ~ 1)
## Each participant has a unique patterh of the thresholds
random0 = list(list(group = ~ id, gamma = ~ 1))
model0 = make_stan_model(random0, model = 'ordinal', gamma_link = link)
sdata0 = make_stan_data(adata0, fixed0, random0, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit0 = stan(model_code = model0, data = sdata0,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit0, '~/windows/temp/ordinal_fit_0.rds')
}else{
    fit0 = readRDS('~/windows/temp/ordinal_fit_0.rds')
}
```

```{r echo=T,results="show"}              
print(fit0, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
## neff and Rhat look good
```

The neff and Rhat columns look good. Below we translate between the
gamma and the criteria vectors (fixed effects = sample average):

```{r echo=T,results="show"}                
s0 = as.data.frame(fit0)
crit0 = gamma_to_crit(s0, gamma_link = link)
```

Are the sample average thesholds different in the two models? Off
course they are, since the models are non-linear and averaging over
anything that affects any parameter of such models is a bad idea:

```{r echo=T,results="show"}                
crit_sm = as.data.frame(t(apply(crit, 2, sumfun)))
crit0_sm = as.data.frame(t(apply(crit0, 2, sumfun)))
crit_sm$model = 'fixed thresholds'
crit0_sm$model = "random thresholds"
crit_sm$threshold = 1:7 - .1
crit0_sm$threshold = 1:7 + .1
df = rbind(crit_sm, crit0_sm)
names(df)[1:3] = c('lo', 'mean', 'hi')
ggplot(df, aes(threshold, mean, group = model, color = model)) + geom_point() + geom_line() + geom_errorbar(aes(ymin = lo, ymax = hi))
```

The point and interval estimates of the thresholds based on the
thresholds-as-"item-parameters" model are severely biased. In
particular, the uncertainty in the estimates is severely
underestimated in this model. Was it obvious to you that the model
with no threshold random effects is such a bad idea?

Now we will fit a model which is similar to an SDT model, but there
will be two important differences. We will not assume that the answer
"signal" cannot be less likely for the stimulus "signal" than for the
stimulus "noise", i.e., we will not assume that d' must be
non-negative, and we will *not* assume that the pattern of the
thresholds is the same for "noise" and "signal" stimuli. If you know
your Signal Detection Theory well, this may seem to you like a
perverse idea. If that is the case, please bear with me.

In this model each participant may have two different sets of
thresholds, one for "noise" and one for "signal" stimuli. Since the
thresholds are represented as random effects, we will account for the
fact that "random" thresholds may be correlated, and they are - if
some threshold is higher (lower) than the sample average, then the
threshold above (below) it may also be higher (lower).

Here we introduce the effect of the stimulus on the gamma / thresholds
vector. We use separate intercepts and slopes parametrization (-1 +
stim), i.e., the thresholds are to be estimated for each stimulus class
instead of estimating them for "noise" stimuli and estimating the
difference in the thresholds between "signal" and "noise".

```{r echo=T,results="show"}                
fixed1 = list(eta = ~ 1, gamma = ~ -1 + stim)
random1 = list(list(group = ~ id, gamma = ~ -1 + stim))
## We have to make sure that stim is a factor for the separate
## intercepts and slopes parametrization to work as intended.
d2$stim = as.factor(d2$stim)
## Note that because for an ordinal model the stimulus variable (here
## the 'stim' variable) is not special, we have to include it in the
## additional variables vector
adata1 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
sdata1 = make_stan_data(adata1, fixed1, random1, model = 'ordinal', gamma_link = link)
model1 = make_stan_model(random1, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit1 = stan(model_code = model1, data = sdata1,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit1, '~/windows/temp/ordinal_fit_1.rds')
}else{
    fit1 = readRDS('~/windows/temp/ordinal_fit_1.rds')
}
## We will not look at the eta parameter since it was fixed at 0
print(fit1, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
## neff and Rhat look good
```

Now we can translate between the gamma and the criteria vectors (fixed
effects = sample average) for each stimulus:

```{r echo=T,results="show"}
s1 = as.data.frame(fit1)
crit1 = gamma_to_crit(s1, 1, gamma_link = link)
crit2 = gamma_to_crit(s1, 1, beta_index = 2, gamma_link = link)
sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
crit1_sm = apply(crit1, 2, sumfun)
crit2_sm = apply(crit2, 2, sumfun)
```

The criteria for the stimulus class "noise" look like this:

```{r echo=T,results="show"}
round(t(crit1_sm), 2)
```

and the criteria for the stimulus class "signal" look like this:

```{r echo=T,results="show"}
round(t(crit2_sm), 2)
```

Note the difference between the main criterion (row 4). This
difference approximates d'. Even though, as we will later see, this
model does not correctly account for the effect of the stimulus, we can
still recover this effect pretty well, because d' is equivalent to a
constant shift of all the thresholds in one predefined
direction. Shifting all the criteria to the right is equivalent to
shifting the underlying distribution of "latent values" (or evidence)
to the left, so we have to subtract the average position of the
criteria for the "signal" stimuli from the average position of the
criteria for the "noise" stimuli. Here is our first attempt:

```{r echo=T,results="show"}
dprim1 = sumfun(s1[, 'gamma_fixed[4,1]'] - s1[, 'gamma_fixed[4,2]'])
round(dprim1, 2)
```

We will later see that this is our sample average *d'*, but it also
isn't.

In an SDT model we assume that the stimulus class affects *only* the
d' parameter, and that both criteria (gamma) and *d'* may differ
between the participants (or items, etc.). In bhsdtr d' is modelled as
log(d') = delta, which forces it to be non-negative. This is an
important assumption of Signal Detection Theory - if this assumption
is violated, for example, because of the response reversal in some
participants or conditions, then an SDT model is not valid and a
non-trivial generalization of SDT should be used instead (see the
bhsdtr preprint for a more detailed explanation and some
references). Otherwise, this is the same general hierarchical ordinal
regression model.

```{r echo=T,results="show"}   
fixed2 = list(delta = ~ 1, gamma = ~ 1)
random2 = list(list(group = ~ id, delta = ~ 1, gamma = ~ 1))
```

Because of the importance of the stimulus class ("noise", or 1, and
"signal", or 2) in SDT models we have to indicate which variable
represents this information ( aggregate_responses(..., stimulus =
'stim', ...) ).
